---
title: "Project 2"
author: "Mashmuma"
date: "4/28/2020"
output: 
  html_document:
    keep_md: true
---

#setwd("C:/Data Science Specialization/Reproducible Research/Quiz/Peer Assessments/2/")
## Synopsis
The National Weather Service Storm data is read and analyzed in order to determine the types of storm events that are the most harmful to population health as well as the highest economic consequences. The data shows that the biggest source of property damage in the United States comes from Tornados and is followed by Floods and Flash Flood sources. The highest contributor to crop damage in the United States comes from hail and is followed by flood and flash floods. The highest source of injuries and fatalities from weather events both come from tornados. The second highest fatality source comes from flood sources and is followed by lightning then wind. The second largest source of injuries comes from wind followed by lightning sources.  

## Data Processing
The following is the reading and initialization of the dataset:
```{r}
    #Read and initialize library and data
    library(ggplot2)
    library(grid)
    library(gridExtra)
    library('dplyr')
    library('data.table')
    
   url = "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(url, dest = "storm.bz2")
  data = read.csv(bzfile("storm.bz2"))
```

At first, to address the first question of weather events that are most harmful to population, we look at total number of injuries and fatalities by different weather events.

```{r}
library(plyr)
injuries = ddply(data, .(EVTYPE), summarize, sum.injuries = sum(INJURIES,na.rm=TRUE))
injuries = injuries[order(injuries$sum.injuries, decreasing = TRUE), ]
```

Now let’s look at the 5 most harmful weather events.
```{r}
head(injuries, 5)
```

We see that tornado is the most harmful event with injuries of more than 91 thousands. This can be represented in the below figure:

```{r}
ggplot(injuries[1:6, ], aes(EVTYPE, sum.injuries, fill = EVTYPE,alpha=0.5)) + geom_bar(stat = "identity") + 
  xlab("Event Type") + ylab("Number of Injuries") + ggtitle("Injuries by Event type") + coord_flip()
```

the 5 most fatalities events.
```{r}
fatalities = ddply(data, .(EVTYPE), summarize, sum = sum(FATALITIES))
fatalities = fatalities[order(fatalities$sum, decreasing = TRUE), ]
head(fatalities, 5)
```

We see that it is tornado again with fatalities of more than 5 thousands followed by excessive heat causing close to 2 thousands fatalities. We again provide a figure below to give a clear picture in a succint manner.
```{r}
ggplot(fatalities[1:6, ], aes(EVTYPE, sum, fill=EVTYPE,alpha=0.3)) + geom_bar(stat = "identity") + 
  xlab("Event Type") + ylab("Number of Fatalities") + ggtitle("Fatalities by Event type") + coord_flip()
```

economic consequences, we will investigate property damagea followed by crop damage and then total damage. Let’s focus on property damage first. We start by looking at various exponents for PROPDMGEXP.
```{r}
unique(data$PROPDMGEXP)
```

As some have lower character, we convert them to upper character. Also we
replace symbols other than character of numeric values to 0.
```{r}
data$PROPDMGEXP <- toupper(data$PROPDMGEXP)
data$PROPDMGEXP[data$PROPDMGEXP %in% c("", "+", "-", "?")] = "0"
```
As PROPDMGEXP stands for the power of 10, we convert ‘B’ standing for billions to 9, ‘M’ standing for millions to 6, ‘K’ standing for thousands to 3 and ‘H’ for hundreds to 2.

```{r}
data$PROPDMGEXP[data$PROPDMGEXP %in% c("B")] = "9"
data$PROPDMGEXP[data$PROPDMGEXP %in% c("M")] = "6"
data$PROPDMGEXP[data$PROPDMGEXP %in% c("K")] = "3"
data$PROPDMGEXP[data$PROPDMGEXP %in% c("H")] = "2"
```
Now we get the full property damage by converting PROPDMGEXP to numeric values and calculating total damage by multiplying the damage by the corresponding exponent.

```{r}
data$PROPDMGEXP <- 10^(as.numeric(data$PROPDMGEXP))
damage.property = data$PROPDMG * data$PROPDMGEXP
data=as.data.frame(cbind(data,damage.property))
```
Now we make a new dataset of property damage arranged according to events type and look at the first 6 major events in terms of economic loss.

```{r}
Damage.property = ddply(data, .(EVTYPE), summarize, damage.property = sum(damage.property, na.rm = TRUE))
# Sort the Damage dataset
Damage.property = Damage.property[order(Damage.property$damage.property, decreasing = T), ]
# Show the first 6 most damaging types
head(Damage.property)
```

We see that Flood is the major damaging event for housing in terms of economic loss with a total amount of more than 144 billion. This is followed by hurricane/typhoon and tornado.

Now we will look at which event is most devastating economically for crops. As with the economic computation, we take the similar steps and look at the most damaging event for crops in terms of economic loss.

Let’s have a look at various exponents for CROPDMGEXP.
```{r}
unique(data$CROPDMGEXP)
```

As two levels have lower characters, we convert them to upper character. Also we replace symbols other than character of numeric values to 0.

```{r}
data$CROPDMGEXP <- toupper(data$CROPDMGEXP)
data$CROPDMGEXP[data$CROPDMGEXP %in% c("", "?")] = "0"
```
As PROPDMGEXP stands for the power of 10, we convert ‘B’ standing for billions to 9, ‘M’ standing for millions to 6, ‘K’ standing for thousands to 3 and ‘H’ for hundreds to 2.

```{r}
data$CROPDMGEXP[data$CROPDMGEXP %in% c("B")] = "9"
data$CROPDMGEXP[data$CROPDMGEXP %in% c("M")] = "6"
data$CROPDMGEXP[data$CROPDMGEXP %in% c("K")] = "3"
data$CROPDMGEXP[data$CROPDMGEXP %in% c("H")] = "2"
```
Now we get the full crop damage by converting PROPDMGEXP to numeric values and calculating total damage by multiplying the damage by the corresponding exponent.

```{r}
data$CROPDMGEXP <- 10^(as.numeric(data$CROPDMGEXP))
damage.crop = data$CROPDMG * data$CROPDMGEXP
data=as.data.frame(cbind(data,damage.crop))
```

Now we make a new dataset of crop damage arranged according to events type and look at the first 6 major events in terms of economic loss.

```{r}
Damage.crop = ddply(data, .(EVTYPE), summarize, damage.crop = sum(damage.crop, na.rm = TRUE))
# Sort the Damage.crop dataset
Damage.crop = Damage.crop[order(Damage.crop$damage.crop, decreasing = T), ]
# Show the first 6 most damaging types
head(Damage.crop)
```

Let’s also look at a chart to have a quick look at these figures.

```{r}
ggplot(Damage.crop[1:6, ], aes(EVTYPE, damage.crop, fill = EVTYPE, alpha=0.5)) + geom_bar(stat = "identity") + 
  xlab("Event Type") + ylab("Total damages") + ggtitle("Total damages by Event type") + coord_flip()

```

We see that drought is the worst factor for agriculture causing more than 13 billion dollars. This is followed by flood causing more than 5 billion dollars.

Now if we want to look at the econmoic losses at aggregate, we need to add the losses from property and crop and then look at which type is most devastating, flood or drought.

Let’s compute total damage first and combine it to data.Then we just need to segregate losses according to event types.

```{r}
total.damage = damage.property + damage.crop
data=as.data.frame(cbind(data,total.damage))
Damage.total = ddply(data, .(EVTYPE), summarize, damage.total = sum(total.damage, na.rm = TRUE))
# Sort the Damage.crop dataset
Damage.total = Damage.total[order(Damage.total$damage.total, decreasing = T), ]
```
Let’s have a look at first 6 most damaging types

```{r}
head(Damage.total)
```
We see that it is flood with a whooping loss of more than 150 billions followed by hurricane/typhoon with an estimate of more than 71 billions.In terms of total losses, drought–main economic loss event is not even among the loss inducing six events.

## Results
It is evident from the exploratory analysis presented here that flood is the most exacerbating factor for economic loss while tornado is for population health. If agriculture is the main concern, then drought may be the most concerning factor for economy. But for the economy in general, flood becomes the main loss factor. Concerning human health, more priorities naturally will go towards addressing tornado as it claims most lives and causes injuries.